Constructing Quotient Inductive-Inductive Types

### 2018-10-10 paper.tex
third reviewer typos

The previous $\Con$-$\Ty$ example is represented by the context below. We only present here a prefix of the definition which suffices to demonstrate inductive-inductive dependency.
\begin{alignat*}{5}
  & \Theta:\equiv\big(\mathit{Con}:\U,\mathit{Ty}:\mathit{Con}\Ra\U,\cdot:\El\,\mathit{Con},\rhd:(\mathit{\Gamma}:\mathit{Con})\Ra \mathit{Ty}\app\mathit{\Gamma}\Ra\El\,\mathit{Con},...\big)
\end{alignat*}
% U:(\mathit{\Gamma}:\mathit{Con})\Ra\mathit{Ty}\app\mathit{\Gamma}
Algebras for this signature consist of a set, a family of sets over
it, and operators which construct elements of these.
\begin{alignat*}{5}
  & \Theta^\A\equiv(C:\Set)\times(T:C\ra\Set)\times(e:C)\times(f:(\mathit{\gamma}:C)\ra T\,\mathit{\gamma}\ra C)\times...
\end{alignat*}
The initial algebra now has two sorts: one is given by terms of
type $\mathit{Con}$, the other takes a $\mathit{Con}$-term as an input
and outputs the type of terms of type $\mathit{El}$ at the
$\mathit{Con}$-term.
\[
\con_\Theta \equiv \big(\Tm\,\Phi\,(\El\,\mathit{Con}),\, \lambda t.\Tm\,\Phi\,(\El\,(\mathit{Ty}\app t)),\, \cdot,\, \lambda t\,r.\rhd\app t\app r,...\big):\Theta^\A
\]
An algebra homomorphism is given by a function between the
$\mathit{Con}$ components and a function between the $\mathit{Ty}$
components which refers to the first function (this phenomenon is
called recursion-recursion in \cite{forsberg-phd} as an analogue to
induction-induction).
\begin{alignat*}{5}
  & \Theta^\M\,(C,T,e,f,...)\,(C',T',e',f',...)\equiv\, && (C^M:C\ra C')\times(T^M:(\gamma:C)\ra T\,\gamma\ra T'\,(C^M\,\gamma))\,\times \\
  & && (e^M:C^M\,e=e')\,\times \\
  & && (f^M:(\gamma:C)(\alpha:T\,\gamma)\ra C^M\,(f\,\gamma\,\alpha)=f'\,(C^M\,\gamma)\,(T^M\,\gamma\,\alpha))\,\times \\
  & && ...
\end{alignat*}
The first function in the recursor invokes the standard interpretation
on its input, the second one invokes it on its second input. We
abbreviate $(C,T,e,f,...)$ by $\gamma$.
\begin{alignat*}{5}
  & \rec_\Phi\,\gamma \equiv (\lambda t.t^\A\,\gamma,\, \lambda t\,r.r^\A\,\gamma,\, \refl_{e},\,\lambda t\,r.\refl_{f\,(t^\A\,\gamma)\,(r^\A\,\gamma)},...):\Theta^\A
\end{alignat*}
A displayed algebra over an algebra $(C,T,e,f,...)$ consists of a
family over $C$ and a family over $T\,\gamma$ which is also indexed
over the first family.
\begin{alignat*}{5}
  & \Theta^\D\,(C,T,e,f,...)\equiv\, && (C^D:C\ra \Set)\times(T^D:(\gamma:C)\ra C^D\,\gamma\ra T\,\gamma\ra \Set)\,\times \\
  & && (e^D:C^D\,e)\,\times \\
  & && (f^D:(\gamma:C)(\gamma^D:C^D\,\gamma)(\alpha:T\,\gamma)(\alpha^D:T^D\,\gamma\,\gamma^D\,\alpha)\ra C^D\,(f\,\gamma\,\alpha))\times...
\end{alignat*}
As in the case of the homomorphisms, the second function in a section
refers to the first one.
\begin{alignat*}{5}
  & \Theta^\S\,(C,T,e,f,...)\,(C^D,T^D,e^D,f^D,...)\,\equiv \\
  & \hspace{2em} (C^S:(\gamma:C)\ra C^D\,\gamma)\times(T^S:(\gamma:C)(\alpha:T\,\gamma)\ra T^D\,\gamma\,(C^S\,\gamma)\,\alpha)\,\times \\
  & \hspace{2em} (e^S:C^S\,e=e^D)\,\times \\
  & \hspace{2em} (f^S:(\gamma:C)(\alpha:T\,\gamma)\ra C^S\,(f\,\gamma\,\alpha)=f^D\,\gamma\,(C^S\,\gamma)\,\alpha\,(T^S\,\gamma\,\alpha))\times ...
\end{alignat*}
The eliminator is given analogously to the recursor, but using
$\blank^\D$ and $\blank^\C$ instead of $\blank^\A$.

@@

In the following table we summarize the operations which we define in
the Sections \ref{sec:algebras}--\ref{sec:eliminator}. $\Gamma$
denotes a signature. The full definitions of these operations are
given in Appendix \ref{sec:app}.

@@

Quotient types as in \cite{hofmann95extensional} are a precursor to
the current development. More recently, an increasing amount of
research is concerned with higher inductive types (HITs), motivated by
their use cases in homotopy type theory \cite{HoTTbook}. The theory of
HITs is relevant to the current work as it also has to describe
signatures with equality constructors and algebras with
equalities. \cite{niels} define an external syntactic scheme for
higher inductive types with only 0-constructors and compute the types
of elimination principles. In \cite{nielsmsc} a semantics is given for
the same class of HITs but with no recursive equality
constructors. Sojakova \cite{sojakova} defines a subset of HITs
called W-suspensions by an internal coding scheme similar to
W-types. She proves that the induction principle is equivalent to
homotopy initiality. Dybjer and Moeneclaey define a syntactic scheme
for a some HITs
and show their existence in a groupoid model
\cite{moeneclaey}. In \cite{AwodeyFS18}, impredicative encodings of a
class of higher inductive types are presented, and we think that this
approach could also be modelled within our framework once we assume an
impredicative universe.

Lumsdaine and Shulman give a general specification of
models of type theory supporting higher inductive types
\cite{lumsdaineShulman}. They introduce the notion of cell monad with
parameters and characterise the class of models which have intial
algebras for a cell monad with parameters. \cite{cubicalhits} develop
semantics for several HITs (sphere, torus, suspensions, truncations,
pushouts) in certain presheaf toposes, and extends the syntax of
cubical type theory \cite{ctt} with these HITs.

@@

We also assume the existence of one QIIT, namely the syntax of the
type theory of signatures (Section \ref{sec:signatures}). In our
formalisation we postulate its constructors and dependent elimination
principle adding the computation rules as rewrite rules. This QIIT has
equality constructors for the substitution calculus (see the next
section). In this paper, we treat these equalities as definitional (we
don't write transports along them). Analogously, we add rewrite rules
for them in the formalisation.

@@

The universe is interpreted by $\U^\C\,\nu\,a :\equiv
\Tm\,\Omega\,(\El\,a)$ which results in the sorts being modelled by
terms in the initial algebra. For $\El$, we coerce along the equality
$a^\C\,\nu : \Tm\,\Omega\,(\El\,a) = a^\A\,(\Gamma^\C\,\nu)$, and thus
$(\El\,a)^\A\,\nu\,t :\equiv \coe\,(a^\C\,\nu)\,t$. For $\Pi\,a\,B$
types, we define a function which coerces the input along $a^\C$ and
uses $B^\C$ to produce the result in the initial algebra:
\[
(\Pi\,a\,B)^\C\,\nu\,t :\equiv \lambda\alpha.B^\C\,\big(\nu,\coe\,({a^\C\,\nu^{-1}})\,\alpha\big)\,\big(t\app \coe\,({a^\C\,\nu^{-1}})\,\alpha\big)
\]
Equalities of terms and substitutions are trivial in the $\blank^\C$
interpretation, as they are equalities between equalities and these
can be just given by $\UIP$.

@@

Given two $\Gamma$-algebras $\gamma,\gamma':\Gamma^\A$, we define the
notion of homomorphism between them using a variant of the logical
relation interpretation for dependent types
\cite{bernardy12parametricity,atkey}. Contexts become binary
relations, types become heterogeneous binary relations indexed over a
relation for a context, and substitutions and terms are interpreted as
the fundamental theorems for the logical relation.
\begin{alignat*}{5}
  & (\Gamma:\Con)^\M && : \Gamma^\A\ra\Gamma^\A\ra\Set \\
  & (A:\Ty\,\Gamma)^\M && : \Gamma^\M\,\gamma^{\mathit{0}}\,\gamma^{\mathit{1}}\ra A^\A\,\gamma^{\mathit{0}}\ra A^\A\,\gamma^{\mathit{1}}\ra\Set \\
  & (\sigma:\Tms\,\Gamma\,\Delta)^\M && : \Gamma^\M\,\gamma^{\mathit{0}}\,\gamma^{\mathit{1}}\ra\Delta^\M\,(\sigma^\A\,\gamma^{\mathit{0}})\,(\sigma^\A\,\gamma^{\mathit{1}}) \\
  & (t:\Tm\,\Gamma\,A)^\M && : (\gamma^M:\Gamma^\M\,\gamma^{\mathit{0}}\,\gamma^{\mathit{1}})\ra A^\M\,\gamma^M\,(t^\A\,\gamma^{\mathit{0}})\,(t^\A\,\gamma^{\mathit{1}})
\end{alignat*}
However, the usual logical relation interpretation does not produce
homomorphisms. The motivation of Reynolds \cite{reynolds83abstraction}
to replace homomorphisms with logical relations was that homomorphisms
do not work for higher order functions. In our case there are no
higher-order functions because the function space $\Pi$ is strictly
positive. Thus, we are able to interpret the universe by function
space instead of relation space. Similarly, the relation for $\El$ is
the graph of the corresponding function. Below we list the
differences: the left hand side ($\blank^{\M'}$) is the usual logical
relation interpretation, the right hand side is the one we use.
\begin{alignat*}{10}
  & \U^{\M'}\,\gamma^{M'}\,T^{\mathit{0}}\,T^{\mathit{1}} && :\equiv T^{\mathit{0}}\ra T^{\mathit{1}}\ra\U  && \U^\M\,\gamma^M\,T^{\mathit{0}}\,T^{\mathit{1}} && :\equiv T^{\mathit{0}}\ra T^{\mathit{1}} \\
  & (\El\,a)^{\M'}\,\gamma^{M'}\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}} && :\equiv a^{\M'}\,\gamma^{M'}\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}} && (\El\,a)^\M\,\gamma^M\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}} && :\equiv a^\M\,\gamma^M\,\alpha^{\mathit{0}} = \alpha^{\mathit{1}} \\
  & (\Pi\,a\,B)^{\M'}\,\gamma^{M'}\,f^{\mathit{0}}\,f^{\mathit{1}} && :\equiv (\alpha^{M'}:a^{M'}\,\gamma^{M'}\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}})\ra\hspace{2em}  && (\Pi\,a\,B)^\M\,\gamma^M\,f^{\mathit{0}}\,f^{\mathit{1}} && :\equiv (\alpha^{\mathit{0}}:a^\A\,\gamma^{\mathit{0}})\ra \\
  &  && \hspace{1.5em} B^{\M'}\,(\gamma^{M'}, \alpha^{M'})\,(f^{\mathit{0}}\,\alpha^{\mathit{0}}) &&  && \hspace{2em} B^\M\,(\gamma^M, \refl)\,(f^{\mathit{0}}\,\alpha^{\mathit{0}}) \\
  &  && \hspace{1.5em} \hphantom{B^{\M'}\,(\gamma^{M'}, \alpha^{M'})\,} (f^{\mathit{1}}\,\alpha^{\mathit{1}}) &&  && \hspace{3.4em}(f^{\mathit{1}}\,(a^\M\,\gamma^M\,\alpha^{\mathit{0}})) \\
  & (t\app u)^{\M'}\,\gamma^{M'} && :\equiv t^{\M'}\,(u^{\M'}\,\gamma^{M'}) && (t\app u)^\M\,\gamma^M && :\equiv \J\,(t^\M\,\gamma^M\,(u^\A\,\gamma))\,(u^\M\,\gamma^M)
\end{alignat*}
For $\Pi$ types, we could use the usual interpretation, but we have a
better choice. E.g. for the successor constructor of the natural
numbers, the original formulation would give the condition
$(n^{\mathit{0}}:Nat^{\mathit{1}})(n^{\mathit{1}}:Nat^{\mathit{1}})(n^{M}:Nat^M\,n^{\mathit{0}}
= n^{\mathit{1}})\ra
Nat^M\,(\mathit{suc}^{\mathit{0}}\,n^{\mathit{0}}) =
\mathit{suc^{\mathit{1}}}\, n^{\mathit{1}}$. The right hand side
variant strictifies this and results in
$(n^{\mathit{0}}:Nat^{\mathit{0}})\ra
Nat^M\,(\mathit{suc}^{\mathit{1}}\,n^{\mathit{0}}) =
\mathit{suc}^{\mathit{1}}\,(Nat^M\,n^{\mathit{0}})$. The price we have
to pay is that interpreting application requires usage of $\J$.

$\blank^\M$ is constant $\top$ on the identity type, homomorphisms
don't state any condition between identity proofs in different
algebras as we have UIP.

@@

$\U^\R$ is given using the standard interpretation $\blank^\A$. Given
$\nu:\Tms\,\Omega\,\Gamma$ and $a:\Tm\,\Omega\,\U$ we need something
of type $(\alpha:a^\A\,\con)\ra a^\A\,\omega$. We know by $a^\C\,\id$
that $a^\C\con = \Tm\,\Omega\,(\El\,a)$, so coercing $\alpha$ along
this and applying the standard interpretation we get
$(\coe\,({a^\C\,\id}^{-1})\,\alpha)^\A\,\omega : a^\A\,\omega$.

@@

In the previous sections, we have given a part of the semantics of QIITs, in order to make precise notions of algebras and dependent elimination. However, much is still missing:

\begin{itemize}
\item A \emph{category} of algebras and homomorphisms, with the assorted category operations and laws.
\item A proof that the properties of being initial and having dependent elimination are equivalent.
\end{itemize}

@@

We start by defining $\Con$ as the type of $\mathbb{N}$-algebras,
$\Ty$ as displayed $\mathbb{N}$-algebras, $\Tms$ as
$\mathbb{N}$-homomorphisms and $\Ty$ as displayed
$\mathbb{N}$-algebras, using the same definitions as in section
\ref{sec:naturals}.

@@

This concludes the CwF$^\K_{\Eq}$ model of the theory of
signatures. Now, it follows that for each signature, there is a
CwF$^\K_{\Eq}$ of algebras, and thus dependent elimination is
equivalent to unique recursion by Section \ref{sec:initind}. Since for
each signature we have constructed an algebra with elimination in
Section \ref{sec:algebras} and Section \ref{sec:eliminator}, it follows
that the constructed algebras are also initial.

>>>>

The previous $\Con$-$\Ty$ example is represented by the context below. We only present here a prefix of the definition which suffices to demonstrate inductive-inductive dependency.
\begin{alignat*}{5}
  & \Theta:\equiv\big(\mathit{Con}:\U,\mathit{Ty}:\mathit{Con}\Ra\U,\cdot:\El\,\mathit{Con},\rhd:(\mathit{\Gamma}:\mathit{Con})\Ra \mathit{Ty}\app\mathit{\Gamma}\Ra\El\,\mathit{Con},...\big)
\end{alignat*}
% U:(\mathit{\Gamma}:\mathit{Con})\Ra\mathit{Ty}\app\mathit{\Gamma}
Algebras for this signature consist of a set, a family of sets over
it, and operators which construct elements of these.
\begin{alignat*}{5}
  & \Theta^\A\equiv(C:\Set)\times(T:C\ra\Set)\times(e:C)\times(f:(\mathit{\gamma}:C)\ra T\,\mathit{\gamma}\ra C)\times...
\end{alignat*}
The initial algebra now has two sorts: one is given by terms of
type $\mathit{Con}$, the other takes a $\mathit{Con}$-term as an input
and outputs the type of terms of type $\mathit{El}$ at the
$\mathit{Con}$-term.
\[
\con_\Theta \equiv \big(\Tm\,\Theta\,(\El\,\mathit{Con}),\, \lambda t.\Tm\,\Theta\,(\El\,(\mathit{Ty}\app t)),\, \cdot,\, \lambda t\,r.\rhd\app t\app r,...\big):\Theta^\A
\]
An algebra homomorphism is given by a function between the
$\mathit{Con}$ components and a function between the $\mathit{Ty}$
components which refers to the first function (this phenomenon is
called recursion-recursion in \cite{forsberg-phd} as an analogue to
induction-induction).
\begin{alignat*}{5}
  & \Theta^\M\,(C,T,e,f,...)\,(C',T',e',f',...)\equiv\, && (C^M:C\ra C')\times(T^M:(\gamma:C)\ra T\,\gamma\ra T'\,(C^M\,\gamma))\,\times \\
  & && (e^M:C^M\,e=e')\,\times \\
  & && (f^M:(\gamma:C)(\alpha:T\,\gamma)\ra C^M\,(f\,\gamma\,\alpha)=f'\,(C^M\,\gamma)\,(T^M\,\gamma\,\alpha))\,\times \\
  & && ...
\end{alignat*}
The first function in the recursor invokes the standard interpretation
on its input, the second one invokes it on its second input. We
abbreviate $(C,T,e,f,...)$ by $\gamma$.
\begin{alignat*}{5}
  & \rec_\Phi\,\gamma \equiv (\lambda t.t^\A\,\gamma,\, \lambda t\,r.r^\A\,\gamma,\, \refl_{e},\,\lambda t\,r.\refl_{f\,(t^\A\,\gamma)\,(r^\A\,\gamma)},...):\Theta^\A
\end{alignat*}
A displayed algebra over an algebra $(C,T,e,f,...)$ consists of a
family over $C$ and a family over $T\,\gamma$ which is also indexed
over the first family.
\begin{alignat*}{5}
  & \Theta^\D\,(C,T,e,f,...)\equiv\, && (C^D:C\ra \Set)\times(T^D:(\gamma:C)\ra C^D\,\gamma\ra T\,\gamma\ra \Set)\,\times \\
  & && (e^D:C^D\,e)\,\times \\
  & && (f^D:(\gamma:C)(\gamma^D:C^D\,\gamma)(\alpha:T\,\gamma)(\alpha^D:T^D\,\gamma\,\gamma^D\,\alpha)\ra C^D\,(f\,\gamma\,\alpha))\times...
\end{alignat*}
As in the case of the homomorphisms, the second function in a section
refers to the first one.
\begin{alignat*}{5}
  & \Theta^\S\,(C,T,e,f,...)\,(C^D,T^D,e^D,f^D,...)\,\equiv \\
  & \hspace{2em} (C^S:(\gamma:C)\ra C^D\,\gamma)\times(T^S:(\gamma:C)(\alpha:T\,\gamma)\ra T^D\,\gamma\,(C^S\,\gamma)\,\alpha)\,\times \\
  & \hspace{2em} (e^S:C^S\,e=e^D)\,\times \\
  & \hspace{2em} (f^S:(\gamma:C)(\alpha:T\,\gamma)\ra C^S\,(f\,\gamma\,\alpha)=f^D\,\gamma\,(C^S\,\gamma)\,\alpha\,(T^S\,\gamma\,\alpha))\times ...
\end{alignat*}
The eliminator is given analogously to the recursor, but using
$\blank^\D$ and $\blank^\C$ instead of $\blank^\A$.

@@

In the following table we summarize the operations which we define in
Sections \ref{sec:algebras}--\ref{sec:eliminator}. $\Gamma$ denotes a
signature. The full definitions of these operations are given in
Appendix \ref{sec:app}.

@@

Quotient types as in \cite{hofmann95extensional} are a precursor to
the current development. More recently, an increasing amount of
research is concerned with higher inductive types (HITs), motivated by
their use cases in homotopy type theory \cite{HoTTbook}. The theory of
HITs is relevant to the current work as it also has to describe
signatures with equality constructors and algebras with
equalities. \cite{niels} define an external syntactic scheme for
higher inductive types with only 0-constructors and compute the types
of elimination principles. In \cite{nielsmsc} a semantics is given for
the same class of HITs but with no recursive equality
constructors. Sojakova \cite{sojakova} defines a subset of HITs called
W-suspensions by an internal coding scheme similar to W-types. She
proves that the induction principle is equivalent to homotopy
initiality. Dybjer and Moeneclaey define a syntactic scheme for some
HITs and show their existence in a groupoid model
\cite{moeneclaey}. In \cite{AwodeyFS18}, impredicative encodings of a
class of higher inductive types are presented, and we think that this
approach could also be modelled within our framework once we assume an
impredicative universe.

Lumsdaine and Shulman give a general specification of
models of type theory supporting higher inductive types
\cite{lumsdaineShulman}. They introduce the notion of cell monad with
parameters and characterise the class of models which have intial
algebras for a cell monad with parameters. \cite{cubicalhits} develop
semantics for several HITs (sphere, torus, suspensions, truncations,
pushouts) in certain presheaf toposes, and extend the syntax of
cubical type theory \cite{ctt} with these HITs.

@@

We also assume the existence of one QIIT, namely the syntax of the
type theory of signatures (Section \ref{sec:signatures}). In our
formalisation we postulate its constructors and dependent elimination
principle, adding the computation rules as rewrite rules. This QIIT has
equality constructors for the substitution calculus (see the next
section). In this paper, we treat these equalities as definitional (we
don't write transports along them). Analogously, we add rewrite rules
for them in the formalisation.

@@

The universe is interpreted by $\U^\C\,\nu\,a :\equiv
\Tm\,\Omega\,(\El\,a)$ which results in the sorts being modelled by
terms in the initial algebra. For $\El$, we coerce along the equality
$a^\C\,\nu : \Tm\,\Omega\,(\El\,a) = a^\A\,(\Gamma^\C\,\nu)$, and thus
$(\El\,a)^\C\,\nu\,t :\equiv \coe\,(a^\C\,\nu)\,t$. For $\Pi\,a\,B$
types, we define a function which coerces the input along $a^\C$ and
uses $B^\C$ to produce the result in the initial algebra:
\[
(\Pi\,a\,B)^\C\,\nu\,t :\equiv \lambda\alpha.B^\C\,\big(\nu,\coe\,({a^\C\,\nu^{-1}})\,\alpha\big)\,\big(t\app \coe\,({a^\C\,\nu^{-1}})\,\alpha\big)
\]
Equalities of terms and substitutions are trivial in the $\blank^\C$
interpretation, as they are equalities between equalities and these
can be just given by $\UIP$.

@@

Given two $\Gamma$-algebras $\gamma,\gamma':\Gamma^\A$, we define the
notion of homomorphism between them using a variant of the logical
relation interpretation for dependent types
\cite{bernardy12parametricity,atkey}. Contexts become binary
relations, types become heterogeneous binary relations indexed over a
relation for a context, and substitutions and terms are interpreted as
the fundamental theorems for the logical relation.
\begin{alignat*}{5}
  & (\Gamma:\Con)^\M && : \Gamma^\A\ra\Gamma^\A\ra\Set \\
  & (A:\Ty\,\Gamma)^\M && : \Gamma^\M\,\gamma^{\mathit{0}}\,\gamma^{\mathit{1}}\ra A^\A\,\gamma^{\mathit{0}}\ra A^\A\,\gamma^{\mathit{1}}\ra\Set \\
  & (\sigma:\Tms\,\Gamma\,\Delta)^\M && : \Gamma^\M\,\gamma^{\mathit{0}}\,\gamma^{\mathit{1}}\ra\Delta^\M\,(\sigma^\A\,\gamma^{\mathit{0}})\,(\sigma^\A\,\gamma^{\mathit{1}}) \\
  & (t:\Tm\,\Gamma\,A)^\M && : (\gamma^M:\Gamma^\M\,\gamma^{\mathit{0}}\,\gamma^{\mathit{1}})\ra A^\M\,\gamma^M\,(t^\A\,\gamma^{\mathit{0}})\,(t^\A\,\gamma^{\mathit{1}})
\end{alignat*}
However, the usual logical relation interpretation does not produce
homomorphisms. The motivation of Reynolds \cite{reynolds83abstraction}
to replace homomorphisms with logical relations was that homomorphisms
do not work for higher order functions. In our case there are no
higher-order functions because the function space $\Pi$ is strictly
positive. Thus, we are able to interpret the universe by function
space instead of relation space. Similarly, the relation for $\El$ is
the graph of the corresponding function. Below we list the
differences: the left hand side ($\blank^{\M'}$) is the usual logical
relation interpretation, the right hand side is the one we use.
\begin{alignat*}{10}
  & \U^{\M'}\,\gamma^{M'}\,T^{\mathit{0}}\,T^{\mathit{1}} && :\equiv T^{\mathit{0}}\ra T^{\mathit{1}}\ra\U  && \U^\M\,\gamma^M\,T^{\mathit{0}}\,T^{\mathit{1}} && :\equiv T^{\mathit{0}}\ra T^{\mathit{1}} \\
  & (\El\,a)^{\M'}\,\gamma^{M'}\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}} && :\equiv a^{\M'}\,\gamma^{M'}\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}} && (\El\,a)^\M\,\gamma^M\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}} && :\equiv a^\M\,\gamma^M\,\alpha^{\mathit{0}} = \alpha^{\mathit{1}} \\
  & (\Pi\,a\,B)^{\M'}\,\gamma^{M'}\,f^{\mathit{0}}\,f^{\mathit{1}} && :\equiv (\alpha^{M'}:a^{M'}\,\gamma^{M'}\,\alpha^{\mathit{0}}\,\alpha^{\mathit{1}})\ra\hspace{2em}  && (\Pi\,a\,B)^\M\,\gamma^M\,f^{\mathit{0}}\,f^{\mathit{1}} && :\equiv (\alpha^{\mathit{0}}:a^\A\,\gamma^{\mathit{0}})\ra \\
  &  && \hspace{1.5em} B^{\M'}\,(\gamma^{M'}, \alpha^{M'})\,(f^{\mathit{0}}\,\alpha^{\mathit{0}}) &&  && \hspace{2em} B^\M\,(\gamma^M, \refl)\,(f^{\mathit{0}}\,\alpha^{\mathit{0}}) \\
  &  && \hspace{1.5em} \hphantom{B^{\M'}\,(\gamma^{M'}, \alpha^{M'})\,} (f^{\mathit{1}}\,\alpha^{\mathit{1}}) &&  && \hspace{3.4em}(f^{\mathit{1}}\,(a^\M\,\gamma^M\,\alpha^{\mathit{0}})) \\
  & (t\app u)^{\M'}\,\gamma^{M'} && :\equiv t^{\M'}\,(u^{\M'}\,\gamma^{M'}) && (t\app u)^\M\,\gamma^M && :\equiv \J\,(t^\M\,\gamma^M\,(u^\A\,\gamma))\,(u^\M\,\gamma^M)
\end{alignat*}
For $\Pi$ types, we could use the usual interpretation, but we have a
better choice. E.g. for the successor constructor of the natural
numbers, the original formulation would give the condition
$(n^{\mathit{0}}:Nat^{\mathit{0}})(n^{\mathit{1}}:Nat^{\mathit{1}})(n^{M}:Nat^M\,n^{\mathit{0}}
= n^{\mathit{1}})\ra
Nat^M\,(\mathit{suc}^{\mathit{0}}\,n^{\mathit{0}}) =
\mathit{suc^{\mathit{1}}}\, n^{\mathit{1}}$. The right hand side
variant strictifies this and results in
$(n^{\mathit{0}}:Nat^{\mathit{0}})\ra
Nat^M\,(\mathit{suc}^{\mathit{1}}\,n^{\mathit{0}}) =
\mathit{suc}^{\mathit{1}}\,(Nat^M\,n^{\mathit{0}})$. The price we have
to pay is that interpreting application requires usage of $\J$.

$\blank^\M$ is constant $\top$ on the identity type; homomorphisms
do not state any conditions between identity proofs in different
algebras because of UIP.

@@

$\U^\R$ is given using the standard interpretation $\blank^\A$. Given
$\nu:\Tms\,\Omega\,\Gamma$ and $a:\Tm\,\Omega\,\U$ we need something
of type $(\alpha:a^\A\,\con)\ra a^\A\,\omega$. We know by $a^\C\,\id$
that $\Tm\,\Omega\,(\El\,a)=a^\A\con$, so coercing $\alpha$ along
this and applying the standard interpretation we get
$(\coe\,({a^\C\,\id}^{-1})\,\alpha)^\A\,\omega : a^\A\,\omega$.

@@

In the previous sections, we have given a part of the semantics of QIITs, in order to make precise notions of algebras and induction. However, much is still missing:

\begin{itemize}
\item A \emph{category} of algebras and homomorphisms, with the assorted category operations and laws.
\item A proof that the properties of being initial and having induction are equivalent.
\end{itemize}

@@

We start by defining $\Con$ as the type of $\mathbb{N}$-algebras,
$\Ty$ as displayed $\mathbb{N}$-algebras, $\Tms$ as
$\mathbb{N}$-homomorphisms and $\Tm$ as displayed
$\mathbb{N}$-algebra sections, using the same definitions as in section
\ref{sec:naturals}.

@@

This concludes the CwF$^\K_{\Eq}$ model of the theory of
signatures. Now, it follows that for each signature, there is a
CwF$^\K_{\Eq}$ of algebras, and thus induction is equivalent to
initiality by Section \ref{sec:initind}. Since for each signature we
have constructed an algebra with induction in Section
\ref{sec:algebras} and Section \ref{sec:eliminator}, it follows that
the constructed algebras are also initial.
